Estimating duration using survival analysis

### 2024-07-31 cis-perfect-testing.tex
improve figure appearance

Uninformative Beta priors on the hazard are problematic in this context.
An uninformative prior would be of the form $\lambda_t \dist \BetaDist(\alpha, \alpha)$.
Commonly, $\alpha$ is chosen to be $0.5$ or $1$ (a uniform prior).
The former, $\alpha = 0.5$, is Jefferys’ prior~\autocite{jeffreysInvariant}, commonly considered an uninformative prior because it is invariant to parameterization~\autocite[53]{gelmanBDA}.
Even though these priors are uninformative on the hazard, they become highly informative on the survival. 
Specifically, they tend to favour shorter survival times (see \cref{perf-test:fig:flat-prior}).
An intuitive explanation of this follows from the fact that priors of this form have an expected value of 0.5.
The expected survival time, $\E \left( \prod_{i=1}^{t-1} (1-\lambda_t) \right)$, is therefore equal to $0.5^{t-1}$, a quantity that declines rapidly.
This prior expresses extreme scepticism in estimates from previous studies, such as \cref{E-ATACCC} and the meta-analysis of \textcite{cevikShedding}.
\begin{figure}
  \centering \includegraphics{cis-perfect-testing/flat-prior}
  \caption[Uninformative priors for the hazard]{
    (A) Density of Beta(1, 1) flat prior and Beta(0.5, 0.5) uninformative priors.
    (B) Prior predictive survival time (mean and 95\% CrI).
    The implied prior on the survival time is short when using either prior on the hazard at each time.
  }
 \label{perf-test:fig:flat-prior}
\end{figure}

Instead, I propose the vague prior Beta(0.1, 1.9), which has mean 0.05 and minimal information.
The amount of information in a Beta distribution is related to the sum of its parameters.
Here, the sum equals 2, the same as the flat prior case.
The central 95\% probability mass of Beta(0.1, 1.9) is 0.00--0.47.
The central estimate, of 0.05, is in line with previous estimates that the median duration in the range 15--20 days~\autocite{cevikShedding}.
This prior gives a very vague prior predictive distribution on the survival time (see \cref{perf-test:fig:vague-prior}).
\begin{figure}
  \centering \includegraphics{cis-perfect-testing/vague-prior}
  \caption[Vague prior for the hazard]{
    (A) Density of Beta(0.1, 1.9) vague prior.
    (B) Prior predictive survival time (mean and 95\% CrI).
    Note the different x-axis to \cref{perf-test:fig:flat-prior}.
  }
  \label{perf-test:fig:vague-prior}
\end{figure}

@@

An RW2 prior encodes that the hazard should be linearly changing with some random changes at each time step.
Specifically, the difference on the logit scale between $\lambda_t$ to $\lambda_{t+1}$ is the difference between $\lambda_{t-1}$ and $\lambda_t$ plus some normally distributed noise.
Formally:
\begin{align}
  \logit\lambda_{t+1}
  &= \logit\lambda_t + (\logit\lambda_t - \logit\lambda_{t-1}) + \sigma \epsilon_t &\text{for $t \geq 2$} \\
  &= 2\logit\lambda_t - \logit\lambda_{t-1} + \sigma \epsilon_t \\
  \epsilon_t &\dist \N(0, 1) &\text{for $t \geq 2$}  \\
  \logit\lambda_2 &= \logit\lambda_1 + \epsilon_1 \\\
  \epsilon_1 &\dist \N(\mu_{\epsilon_1}, \sigma_{\epsilon_1}^2) \\
  \logit \lambda_1 &\dist \N(\mu_{\lambda_1}, \sigma_{\lambda_1}^2) \\
  \sigma &\dist \Exponential(1/\mu_\sigma).
\end{align}
The hyperparameters $\mu_{\lambda_1}$ and $\sigma_{\lambda_1}$ specify the prior on the hazard at time 1; $\mu_{\epsilon_1}$ and $\sigma_{\epsilon_1}$ specify the prior on the initial gradient, $\lambda_2 - \lambda_1$; and $\mu_\sigma$ specifies the smoothness of the random walk.
I use $\mu_{\lambda_1} = -17.5$, $\sigma_{\lambda_1} = 6$, $\mu_\epsilon = 1.09$, $\sigma_\epsilon = 0.03$, and $\mu_\epsilon = 0.1$ based on the analysis in \cref{E-ATACCC} (see \cref{perf-test:fig:rw2-prior}).
\begin{figure}
  \centering \includegraphics{cis-perfect-testing/rw2-prior}
  \caption[RW2 prior for the hazard]{The prior predictive survival time (mean and 95\% central interval) for the RW2 prior specified in the main text. The posterior mean estimate using ATACCC data in \cref{E-ATACCC} is shown for comparison. \label{perf-test:fig:rw2-prior}}
\end{figure}

@@

I first approximate the \cref{E-ATACCC} posterior estimate of the hazard as $\logit{\vec{h}} \dist \MNorm(\vec{\mu}_A, \matr{\Sigma}_A)$ where $\vec{h} = [h_1, \dots, h_{\dmax}]^T$ is the hazard, and $\vec{\mu}_A$ and $\matr{\Sigma}_A$ are the mean and covariance matrix computed from the samples of $\logit{\vec{h}}$ derived using the method of \cref{E-ATACCC:sec:duration}.
Therefore, $\vec{h}$ represents a sample from the approximation of \cref{E-ATACCC}'s posterior estimate of the hazard before increasing the uncertainty to account for the strong model assumptions.
Using a multivariate normal, as opposed to multiple univariate distribution for each $h_t$, preserves the correlation between the hazards.
% The logit transformation maps from the $[0, 1]$ interval to the full real line.
The approximation is very good (see \cref{perf-test:fig:approximate-ATACCC-hazard} and \cref{perf-test:fig:approximate-ATACCC-survival}).
\begin{figure}
  \thisfloatpagestyle{empty}
  % \vspace{-1cm}
  \makebox[\textwidth][c]{\includegraphics[width=.9\paperwidth]{cis-perfect-testing/ataccc-approximation-hazard}}
  \caption[Approximating the ATACCC posterior hazard as a logit-normal]{Comparison of the posterior estimate of the hazard from \cref{E-ATACCC} and its approximation with a logit-normal distribution (kernel density estimate smoothed) for hazards between day 4 and 15. The density for smaller hazards is concentrated very close to 0. \label{perf-test:fig:approximate-ATACCC-hazard}}
\end{figure}
\begin{figure}
  \centering \includegraphics{cis-perfect-testing/ataccc-approximation-survival}
  \caption[Approximating the ATACCC posterior survival]{Comparison of the posterior estimate of the survival from \cref{E-ATACCC} and its approximation with a logit-normal distribution (median and central 95\% interval). The survival is a function of all the hazards up until that point, therefore the correlation between the hazards must be well-approximated to have these agree. \label{perf-test:fig:approximate-ATACCC-survival}}
\end{figure}

@@

\begin{figure}
  \centering \includegraphics{cis-perfect-testing/kt-prior}
  \caption{Function used for $k_t$. \label{perf-test:fig:kt}}
\end{figure}

@@

To form the duration distribution used in the simulation, we combine the two estimates.
The pdf over the first 30 days is proportional to the ATACCC estimate, with the rest proportional to this CIS-based estimate.
Denote by $f_A(t)$ the distribution function estimated in \cref{E-ATACCC} and $f_C(t)$ that from the CIS-based estimates just derived.
Then define:
$$
f_S'(t) = \begin{cases}
	f_A(t) &t \leq 30 \\
	f_C(t) &t > 30
\end{cases}
$$
Then the distribution used in the simulation is the normalised version of this: $f_S(t) = f'_S(t)/\sum_i f_S'(i)$.
These curves and the combined curve are compared in \cref{perf-test:fig:duration-dist}.
Compared to the ATACCC estimate alone, incorporating the CIS portion shifts some probability mass into the tail of the distribution.
Therefore, the survival probability is initially higher but later lower than ATACCC alone.
Individual $i$'s duration of detectability, $D_i$, is then an independent draw from this distribution.
\begin{figure}
  \centering \includegraphics{cis-perfect-testing/input-duration-dists}
  \caption[Comparison of duration distributions]{Comparison of the duration distribution used for the simulation. ATACCC is the posterior mean from \cref{E-ATACCC}'s analysis, CIS is Sarah Walker's analysis, and Combined is the combination of these used for the simulation. \label{perf-test:fig:duration-dist}}
\end{figure}

@@

\begin{figure}
  \centering \includegraphics{cis-perfect-testing/survival-results}
  \caption[Comparison of survival function estimates under different priors]{Comparison of the posterior estimate of the survival distribution (pointwise median and 95\% CrI) when using different priors on the hazard. Results using each of the three different priors discussed in \cref{perf-test:sec:parameters-priors} are shown. y-axis has either natural (left) or log (right) scale. I show only the first 100 days of the distribution for readability. Black line shows the value used in the simulation (see \cref{perf-test:sec:sim-setup}). \label{perf-test:fig:survival-results}}
\end{figure}

\begin{figure}
  \centering \includegraphics{cis-perfect-testing/hazard-results}
  \caption[Comparison of hazard estimates under different priors]{Comparison of the posterior estimate of the hazard (pointwise median and 95\% CrI) when using different priors on the hazard. Results using each of the three different priors discussed in \cref{perf-test:sec:parameters-priors} are shown. Black line shows the value used in the simulation (see \cref{perf-test:sec:sim-setup}). \label{perf-test:fig:hazard-results}}
\end{figure}

>>>>

Uninformative Beta priors on the hazard are problematic in this context.
An uninformative prior would be of the form $\lambda_t \dist \BetaDist(\alpha, \alpha)$.
Commonly, $\alpha$ is chosen to be $0.5$ or $1$ (a uniform prior).
The former, $\alpha = 0.5$, is Jefferys’ prior~\autocite{jeffreysInvariant}, commonly considered an uninformative prior because it is invariant to parameterization~\autocite[53]{gelmanBDA}.
Even though these priors are uninformative on the hazard, they become highly informative on the survival. 
Specifically, they tend to favour shorter survival times (see \cref{perf-test:fig:flat-prior}).
An intuitive explanation of this follows from the fact that priors of this form have an expected value of 0.5.
The expected survival time, $\E \left( \prod_{i=1}^{t-1} (1-\lambda_t) \right)$, is therefore equal to $0.5^{t-1}$, a quantity that declines rapidly.
This prior expresses extreme scepticism in estimates from previous studies, such as \cref{E-ATACCC} and the meta-analysis of \textcite{cevikShedding}.
\begin{figure}
  \makebox[\textwidth][c]{\includegraphics{cis-perfect-testing/flat-prior}}
  \caption[Uninformative priors for the hazard]{
    (A) Density of Beta(1, 1) flat prior and Beta(0.5, 0.5) uninformative priors.
    (B) Prior predictive survival time (mean and 95\% CrI).
    The implied prior on the survival time is short when using either prior on the hazard at each time.
  }
 \label{perf-test:fig:flat-prior}
\end{figure}

Instead, I propose the vague prior Beta(0.1, 1.9), which has mean 0.05 and minimal information.
The amount of information in a Beta distribution is related to the sum of its parameters.
Here, the sum equals 2, the same as the flat prior case.
The central 95\% probability mass of Beta(0.1, 1.9) is 0.00--0.47.
The central estimate, of 0.05, is in line with previous estimates that the median duration in the range 15--20 days~\autocite{cevikShedding}.
This prior gives a very vague prior predictive distribution on the survival time (see \cref{perf-test:fig:vague-prior}).
\begin{figure}
  \makebox[\textwidth][c]{\includegraphics{cis-perfect-testing/vague-prior}}
  \caption[Vague prior for the hazard]{
    (A) Density of Beta(0.1, 1.9) vague prior.
    (B) Prior predictive survival time (mean and 95\% CrI).
    Note the different x-axis to \cref{perf-test:fig:flat-prior}.
  }
  \label{perf-test:fig:vague-prior}
\end{figure}

@@

An RW2 prior encodes that the hazard should be linearly changing with some random changes at each time step.
Specifically, the difference on the logit scale between $\lambda_t$ to $\lambda_{t+1}$ is the difference between $\lambda_{t-1}$ and $\lambda_t$ plus some normally distributed noise.
Formally:
\begin{align}
  \logit\lambda_{t+1}
  &= \logit\lambda_t + (\logit\lambda_t - \logit\lambda_{t-1}) + \sigma \epsilon_t &\text{for $t \geq 2$} \\
  &= 2\logit\lambda_t - \logit\lambda_{t-1} + \sigma \epsilon_t \\
  \epsilon_t &\dist \N(0, 1) &\text{for $t \geq 2$}  \\
  \logit\lambda_2 &= \logit\lambda_1 + \epsilon_1 \\\
  \epsilon_1 &\dist \N(\mu_{\epsilon_1}, \sigma_{\epsilon_1}^2) \\
  \logit \lambda_1 &\dist \N(\mu_{\lambda_1}, \sigma_{\lambda_1}^2) \\
  \sigma &\dist \Exponential(1/\mu_\sigma).
\end{align}
The hyperparameters $\mu_{\lambda_1}$ and $\sigma_{\lambda_1}$ specify the prior on the hazard at time 1; $\mu_{\epsilon_1}$ and $\sigma_{\epsilon_1}$ specify the prior on the initial gradient, $\lambda_2 - \lambda_1$; and $\mu_\sigma$ specifies the smoothness of the random walk.
I use $\mu_{\lambda_1} = -17.5$, $\sigma_{\lambda_1} = 6$, $\mu_\epsilon = 1.09$, $\sigma_\epsilon = 0.03$, and $\mu_\epsilon = 0.1$ based on the analysis in \cref{E-ATACCC} (see \cref{perf-test:fig:rw2-prior}).
\begin{figure}
  \makebox[\textwidth][c]{\includegraphics{cis-perfect-testing/rw2-prior}}
  \caption[RW2 prior for the hazard]{The prior predictive survival time (mean and 95\% central interval) for the RW2 prior specified in the main text. The posterior mean estimate using ATACCC data in \cref{E-ATACCC} is shown for comparison. \label{perf-test:fig:rw2-prior}}
\end{figure}

@@

I first approximate the \cref{E-ATACCC} posterior estimate of the hazard as $\logit{\vec{h}} \dist \MNorm(\vec{\mu}_A, \matr{\Sigma}_A)$ where $\vec{h} = [h_1, \dots, h_{\dmax}]^T$ is the hazard, and $\vec{\mu}_A$ and $\matr{\Sigma}_A$ are the mean and covariance matrix computed from the samples of $\logit{\vec{h}}$ derived using the method of \cref{E-ATACCC:sec:duration}.
Therefore, $\vec{h}$ represents a sample from the approximation of \cref{E-ATACCC}'s posterior estimate of the hazard before increasing the uncertainty to account for the strong model assumptions.
Using a multivariate normal, as opposed to multiple univariate distribution for each $h_t$, preserves the correlation between the hazards.
% The logit transformation maps from the $[0, 1]$ interval to the full real line.
The approximation is very good (see \cref{perf-test:fig:approximate-ATACCC-hazard} and \cref{perf-test:fig:approximate-ATACCC-survival}).
\begin{figure}
  \thisfloatpagestyle{empty}
  % \vspace{-1cm}
  \makebox[\textwidth][c]{\includegraphics{cis-perfect-testing/ataccc-approximation-hazard}}
  \caption[Approximating the ATACCC posterior hazard as a logit-normal]{Comparison of the posterior estimate of the hazard from \cref{E-ATACCC} and its approximation with a logit-normal distribution (kernel density estimate smoothed) for hazards between day 4 and 15. The density for smaller hazards is concentrated very close to 0. \label{perf-test:fig:approximate-ATACCC-hazard}}
\end{figure}
\begin{figure}
  \makebox[\textwidth][c]{\includegraphics{cis-perfect-testing/ataccc-approximation-survival}}
  \caption[Approximating the ATACCC posterior survival]{Comparison of the posterior estimate of the survival from \cref{E-ATACCC} and its approximation with a logit-normal distribution (median and central 95\% interval). The survival is a function of all the hazards up until that point, therefore the correlation between the hazards must be well-approximated to have these agree. \label{perf-test:fig:approximate-ATACCC-survival}}
\end{figure}

@@

\begin{figure}
  \makebox[\textwidth][c]{\includegraphics{cis-perfect-testing/kt-prior}}
  \caption{Function used for $k_t$. \label{perf-test:fig:kt}}
\end{figure}

@@

To form the duration distribution used in the simulation, we combine the two estimates.
The pdf over the first 30 days is proportional to the ATACCC estimate, with the rest proportional to this CIS-based estimate.
Denote by $f_A(t)$ the distribution function estimated in \cref{E-ATACCC} and $f_C(t)$ that from the CIS-based estimates just derived.
Then define:
$$
f_S'(t) = \begin{cases}
	f_A(t) &t \leq 30 \\
	f_C(t) &t > 30
\end{cases}
$$
Then the distribution used in the simulation is the normalised version of this: $f_S(t) = f'_S(t)/\sum_i f_S'(i)$.
These curves and the combined curve are compared in \cref{perf-test:fig:duration-dist}.
Compared to the ATACCC estimate alone, incorporating the CIS portion shifts some probability mass into the tail of the distribution.
Therefore, the survival probability is initially higher but later lower than ATACCC alone.
Individual $i$'s duration of detectability, $D_i$, is then an independent draw from this distribution.
\begin{figure}
  \makebox[\textwidth][c]{\includegraphics{cis-perfect-testing/input-duration-dists}}
  \caption[Comparison of duration distributions]{Comparison of the duration distribution used for the simulation. ATACCC is the posterior mean from \cref{E-ATACCC}'s analysis, CIS is Sarah Walker's analysis, and Combined is the combination of these used for the simulation. \label{perf-test:fig:duration-dist}}
\end{figure}

@@

\begin{figure}
  \makebox[\textwidth][c]{\includegraphics{cis-perfect-testing/survival-results}}
  \caption[Comparison of survival function estimates under different priors]{Comparison of the posterior estimate of the survival distribution (pointwise median and 95\% CrI) when using different priors on the hazard. Results using each of the three different priors discussed in \cref{perf-test:sec:parameters-priors} are shown. y-axis has either natural (left) or log (right) scale. I show only the first 100 days of the distribution for readability. Black line shows the value used in the simulation (see \cref{perf-test:sec:sim-setup}). \label{perf-test:fig:survival-results}}
\end{figure}

\begin{figure}
  \makebox[\textwidth][c]{\includegraphics{cis-perfect-testing/hazard-results}}
  \caption[Comparison of hazard estimates under different priors]{Comparison of the posterior estimate of the hazard (pointwise median and 95\% CrI) when using different priors on the hazard. Results using each of the three different priors discussed in \cref{perf-test:sec:parameters-priors} are shown. Black line shows the value used in the simulation (see \cref{perf-test:sec:sim-setup}). \label{perf-test:fig:hazard-results}}
\end{figure}
